---
title: 'Schedule'
include-in-header:
  text: <script type="text/javascript" src="highlight_date.js"></script>
---

Please consult [Smith's academic calendar](https://www.smith.edu/academics/academic-calendar).

```{r}
#| message: false
#| echo: false
classes <- c(
  as.Date("2025-09-08") + lubridate::ddays(0:13*7), 
  as.Date("2025-09-10") + lubridate::ddays(0:13*7)
) |>
  sort()

i <- 0
echo_date <- function() {
  i <<- i + 1
  x <- classes[i] |>
    format("%a, %b %e")
  htmltools::div(
    class = "card-header",
    "data-date" = classes[i],
    htmltools::h4(paste0(i, ". ", x))
  )
}
```

### Intro

::: {.grid}

::: {.g-col-6 .card} 
`r echo_date()`

- Introductions
- Meet our project sponsors

::: {.callout-note} 
## Homework

- Complete [project selection form](https://docs.google.com/forms/d/e/1FAIpQLSe434T-1ENJmLezTGaEQ7I_55hcCUmtWgtLhiJwD6KBSkim3A/viewform?usp=preview) by **Tuesday at noon** (log into Smith in your browser)

:::

::: 
::: {.g-col-6 .card} 
`r echo_date()`

::: {.callout-note} 
## Homework

- Elect a product owner for Sprint 1
- Contact your project sponsor!
- Pick a time on Friday/Saturday/Sunday for your 3rd weekly standup, post the schedule to Slack, and pin it
:::

:::

::: {.g-col-6 .card}
`r echo_date()`

- Introduction to Teamwork
    - **[Megan Lyster](https://www.smith.edu/people/megan-lyster)**  
      Assistant Director of the Wurtele Center
- Time permitting:
    - Sprint planning:
        - Brainstorm a [Blue Sky](https://en.wiktionary.org/wiki/blue-sky_thinking) vision
        - Set up sprint board using GitHub Projects


::: {.callout-note} 
## Homework

- Read [Codes of Ethics](ethics_codes.qmd)

:::



::: 
::: {.g-col-6 .card}
`r echo_date()`

- `r fontawesome::fa("video")` [Data security](lectures/dev_data_security.html)
- [Codes of Ethics](ethics_codes.qmd)

::: 
::: {.g-col-6 .card} 
`r echo_date()`

- Projects

::: {.callout-note} 
## Homework

- Read [Critique of Codes of Ethics](ethics_critique.qmd)
:::


::: 
::: {.g-col-6 .card}
`r echo_date()`

- [Critique of Codes of Ethics](ethics_critique.qmd)


::: 
::: {.g-col-6 .card} 
`r echo_date()`

- Projects

::: {.callout-note} 
## Homework
- [Book an appointment for a grade conference](https://calendar.app.google/cWz9rKwUyC92GmfV9)
- **Prepare for Sprint Demo**
    - Watch [this video about sprint demos](https://www.youtube.com/watch?v=InXAS_zRvqQ)
    - Read more about [sprint reviews/demos](https://www.atlassian.com/agile/scrum/sprint-reviews)
    - Prepare for 10 minute demo during next class
:::

::: 
::: {.g-col-6 .card} 
`r echo_date()`

```{r}
#| echo: false
#| message: false
#| eval: false
set.seed(21)
library(tidyverse)
sponsors <- c("chicubs", "honda", "cleveland", "elmhurst", "fhi", "dancedata")
start_time <- as_datetime("2025-02-19 13:20:00") + dminutes(0:5 * 10) 
projects <- tibble::tibble(
  time = format(start_time, "%H:%M"),
  demo = sample(sponsors),
  retro = lag(demo, default = last(demo)),
#  planning = lag(retro, default = last(retro))
)

projects |>
  knitr::kable()
```


- **[Sprint demo](sprints.qmd#sprint-demo) (no more than 10 minutes, *with Ben*)**
- **[Sprint retrospective](sprints.qmd#sprint-retrospective) (*with Megan*)**
    - Watch [this video about sprint retrospectives](https://www.youtube.com/watch?v=dJNCHvqKljU)
    - Read more about [sprint retrospectives](https://www.atlassian.com/agile/scrum/retrospectives)
    - Read more about [sprint retros in Scrum](https://www.scrum.org/resources/what-is-a-sprint-retrospective)
    - Revisit your Group Norms (on your README)


::: {.callout-note} 
## Homework

- Submit **Sprint 1 retrospective** to Moodle (by Friday)
- Complete **Reflection and projected grade after Sprint 1** on Moodle (at least 24 hours before your grade conference!)

:::

::: 

::: {.g-col-6 .card} 
`r echo_date()`

- **Sprint planning**

::: {.callout-note} 
## Homework
- `r fontawesome::fa("video")` [Code review](lectures/dev_code_review.html)
:::

::: 
::: {.g-col-6 .card} 
`r echo_date()`

- Ethical Case Study, Part I


::: 
::: {.g-col-6 .card} 
`r echo_date()`

- Projects

::: 
::: {.g-col-6 .card} 
`r echo_date()`

- Ethical Case Study, Part II

::: 
::: {.g-col-6 .card} 
`r echo_date()`

- Projects

::: 
::: {.g-col-6 .card}
`r echo_date()`

- **Sprint 2 Demo**
- **Sprint retro (*with Megan*)**
- **Sprint planning**



:::

::: {.g-col-6 .card} 
`r echo_date()`

:::

::: {.g-col-6 .card}
`r echo_date()`



```{r}
#| echo: false
#| message: false
#| eval: false
projects |>
  mutate(
    demo = sample(demo),
    retro = lag(demo, default = last(demo)),
  ) |>
  knitr::kable()
```



::: 

:::

---

### Sprint 3

::: {.grid} 

::: {.g-col-6 .card} 
`r echo_date()`

- Sprint planning

::: {.callout-note} 
## Homework
- **Submit Final paper, 1st draft by Wednesday at midnight**
- Prepare for Sprint demo
:::


::: 
::: {.g-col-6 .card} 
`r echo_date()`

- [Stochastic parrots](ethics_parrots.html)

::: 
::: {.g-col-6 .card} 
`r echo_date()`

- Projects


::: {.callout-note} 
## Homework

- **Record a presentation by Friday, 5 pm**
    - 1st draft of (first part of) presentation!
    - ~5 minutes
    - Make a few slides
    - Not a demo
    - Concentrate on:
        - What is the problem? 
        - What data do you have?
        - What are the keys issues? Why is this hard?

:::

::: 
::: {.g-col-6 .card} 
`r echo_date()`

- [Generative AI](ethics_ai.html)

::: {.callout-note} 
## Homework

- **Submit Second Reflection by Sunday**
- **Schedule a final presentation [with your client](https://twitter.com/BaumerBen/status/993555071916019712)!**
    - Towards the end of the semester
    - It's OK with me if it happens during reading period or finals week, but you have the right to *not* have it during that time. 
    - I want to be there -- I will work around your schedule

:::

::: 
::: {.g-col-6 .card}
`r echo_date()`

- Projects

::: 
::: {.g-col-6 .card}
`r echo_date()`


```{r}
#| echo: false
#| message: false
#| eval: false
set.seed(784)
library(tidyverse)
start_time <- as_datetime("2024-11-21 09:30:00") + dminutes(0:5 * 10) 
projects <- tibble::tibble(
  time = format(start_time, "%H:%M"),
  demo = sample(c("calibrate", "honda", "cleveland", "lbi", "umass", "dancedata")),
  retro = lag(demo, default = last(demo)),
#  planning = lag(retro, default = last(retro))
)

projects |>
  knitr::kable()
```


- **Sprint 3 Demo**
- **Sprint retro (with Megan)**

::: {.callout-note} 
## Homework
- **Submit Ethics topic statement due Monday**
:::

::: 

::: {.g-col-6 .card}
`r echo_date()`

- Projects

::: {.callout-note} 
## Homework
- **[Submit Final paper, 2nd draft by Friday, 5 pm](paper.html)**
:::


::: 

::: {.g-col-6 .card}
`r echo_date()`

- [Draft feedback](paper.html#feedback-from-previous-drafts)

::: 
::: {.g-col-6 .card}
`r echo_date()`

- Projects

::: 


::: {.g-col-6 .card}
`r echo_date()`

- Projects

::: 
::: {.g-col-6 .card}
`r echo_date()`

- Projects

::: 
::: {.g-col-6 .card}
`r echo_date()`


```{r}
#| echo: false
#| message: false
#| eval: false
tibble::tibble(
  time = as_datetime("2024-09-26 09:30:00") + dminutes(0:5 * 5),
  talk = sample(sponsors)
) |>
  mutate(time = format(time, "%H:%M")) |>
  knitr::kable()
```



- Lightning talks (5 minutes, [2 slides](https://docs.google.com/presentation/d/1qMDYzdpfDkVkYT1a4lB8_k_92FkcSrT_wwwbpWOxukw/edit?usp=sharing))
- Course feedback questionnaire

::: {.callout-note} 
## Homework
- **[Final paper due](paper.html)**
- **[Ethics essay](ethics.html) due**
- **Final reflection due**
- **All work due by May 9**
:::

:::

:::
